---
title: "Casos_Dengue"
author: "Margarita Valdes"
date: "2024-06-11"
output: html_document
---

# Cases of Dengue virus in Mexico in 2024

### Loading libraries

```{r libs}
library(tidyverse)
```

### Reading file

Data was downloaded "Bases de datos Dengue en México 2020" from [Health Secretary](https://datos.gob.mx/busca/dataset/bases-de-datos-dengue-en-mexico). I downloaded both .csv files with the dictionaries for data and catalogue for variable keys.

```{r read_data}
(casos <- read_csv(file = "dengue_abierto.csv"))
```
### Filtering data

I selected the columns I am interested in exploring and making a visuals.

```{r select_data}

(selected_variables <- casos %>% 
          select(ID_REGISTRO,
                 SEXO,
                 EDAD_ANOS,
                 ENTIDAD_RES,
                 FECHA_SIGN_SINTOMAS,
                 TIPO_PACIENTE,
                 HEMORRAGICOS,
                 DEFUNCION,
                 DICTAMEN,
                 TOMA_MUESTRA,
                 RESULTADO_PCR,
                 ESTATUS_CASO))

```

### Re-writing keys 

Using the file "Catalogos_Dengue.xlsx" I rename the keys on most variables, to understand the data set, filter, and make visualizations.

```{r rename_vars}
(rename_variables <- selected_variables %>% 
          mutate(SEXO = case_when(SEXO == 1 ~ "Mujer",
                                  SEXO == 2 ~ "Hombre")) %>% 
          mutate(ENTIDAD_RES = case_when(ENTIDAD_RES == "01" ~ "Aguascalientes",
                                    ENTIDAD_RES == "02" ~ "Baja California",
                                    ENTIDAD_RES == "03" ~ "Baja California Sur",
                                    ENTIDAD_RES == "04" ~ "Campeche",
                                    ENTIDAD_RES == "05" ~ "Coahuila",
                                    ENTIDAD_RES == "06" ~ "Colima",
                                    ENTIDAD_RES == "07" ~ "Chiapas",
                                    ENTIDAD_RES == "08" ~ "Chihuahua",
                                    ENTIDAD_RES == "09" ~ "Ciudad de México",
                                    ENTIDAD_RES == "10" ~ "Durango",
                                    ENTIDAD_RES == "11" ~ "Guanajuato",
                                    ENTIDAD_RES == "12" ~ "Guerrero",
                                    ENTIDAD_RES == "13" ~ "Hidalgo",
                                    ENTIDAD_RES == "14" ~ "Jalisco",
                                    ENTIDAD_RES == "15" ~ "México",
                                    ENTIDAD_RES == "16" ~ "Michoacán",
                                    ENTIDAD_RES == "17" ~ "Morelos",
                                    ENTIDAD_RES == "18" ~ "Nayarit",
                                    ENTIDAD_RES == "19" ~ "Nuevo León",
                                    ENTIDAD_RES == "20" ~ "Oaxaca",
                                    ENTIDAD_RES == "21" ~ "Puebla",
                                    ENTIDAD_RES == "22" ~ "Querétaro",
                                    ENTIDAD_RES == "23" ~ "Quintana Roo",
                                    ENTIDAD_RES == "24" ~ "San Luis Potosí",
                                    ENTIDAD_RES == "25" ~ "Sinaloa",
                                    ENTIDAD_RES == "26" ~ "Sonora",
                                    ENTIDAD_RES == "27" ~ "Tabasco",
                                    ENTIDAD_RES == "28" ~ "Tamaulipas",
                                    ENTIDAD_RES == "29" ~ "Tlaxcala",
                                    ENTIDAD_RES == "30" ~ "Veracruz",
                                    ENTIDAD_RES == "31" ~ "Yucatán",
                                    ENTIDAD_RES == "32" ~ "Zacatecas",
                                    ENTIDAD_RES == "33" ~ "Otros paises",
                                    ENTIDAD_RES == "34" ~ "Otros paises",
                                    ENTIDAD_RES == "35" ~ "Otros paises",
                                    .default = as.character(ENTIDAD_RES))) %>% 
           mutate(TIPO_PACIENTE = case_when(TIPO_PACIENTE == 1 ~ "ambulatorio",
                                            TIPO_PACIENTE == 2 ~ "hospitalizado")) %>% 
           mutate(HEMORRAGICOS = case_when(HEMORRAGICOS == 1 ~ "hemorragico",
                                           HEMORRAGICOS == 2 ~ "no hemorragico")) %>% 
           mutate(DEFUNCION = case_when(DEFUNCION == 1 ~ "si",
                                        DEFUNCION == 2 ~ "no")) %>% 
           mutate(DICTAMEN = case_when(DICTAMEN == 1 ~ "dengue",
                                       DICTAMEN == 2 ~ "chikungunya",
                                       DICTAMEN == 3 ~ "negativo",
                                       DICTAMEN == 4 ~ "en estudio",
                                       DICTAMEN == 5 ~ "no aplica")) %>% 
           mutate(TOMA_MUESTRA = case_when(TOMA_MUESTRA == 1 ~ "si",
                                           TOMA_MUESTRA == 2 ~ "no")) %>% 
           mutate(RESULTADO_PCR = case_when(RESULTADO_PCR == 1 ~ "DENV1",
                                            RESULTADO_PCR == 2 ~ "DENV2",
                                            RESULTADO_PCR == 3 ~ "DENV3",
                                            RESULTADO_PCR == 4 ~ "DENV4",
                                            RESULTADO_PCR == 5 ~ "sin serotipo aislado")) %>% 
           mutate(ESTATUS_CASO = case_when(ESTATUS_CASO == 1 ~ "probable",
                                           ESTATUS_CASO == 2 ~ "confirmado",
                                           ESTATUS_CASO == 3 ~ "descartado"))
                   
 )
          

```



### Save

```{r save}

# write_csv(rename_variables,
#           file = "clean_dengue_data.csv")

```

